---
title: "Cardiometabolic risk in NHANES adults"
author: "Okpara Onyedikachi Martins"
date: today
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(gtsummary)
library(broom)

adults <- read_csv("adults_clean.csv") |>
  mutate(
    AgeDecade = str_trim(AgeDecade),
    smoker = factor(smoker, levels = c("Never", "Former", "Current")),
    bmi_cat = factor(bmi_cat, levels = c("Underweight", "Normal", "Overweight", "Obese"))
  )
```

## Methods

We analysed `r nrow(adults)` adults aged 20 years or older from the NHANES sample.
Hypertension was defined as systolic BP of 140 mmHg or higher, or diastolic BP of
90 mmHg or higher, based on measured values only. Obesity was defined as BMI of
30 kg/m² or higher.

## Results

Hypertension prevalence was `r round(100 * mean(adults$hypertension, na.rm = TRUE), 1)`%
and obesity prevalence was `r round(100 * mean(adults$obese, na.rm = TRUE), 1)`%.

### Table 1. Characteristics by diabetes status

```{r table1}
adults |>
  select(Age, Gender, BMI, bmi_cat, BPSysAve, TotChol, PhysActive, smoker, hypertension, Diabetes) |>
  tbl_summary(
    by = Diabetes,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      TotChol ~ "{median} ({p25}, {p75})"
    ),
    label = list(
      BPSysAve ~ "Systolic BP (mmHg)",
      BMI ~ "BMI (kg/m²)",
      bmi_cat ~ "BMI category",
      TotChol ~ "Total cholesterol (mmol/L)"
    ),
    missing = "no"
  ) |>
  add_p() |>
  add_overall() |>
  bold_labels()
```

### Figure 1. Hypertension prevalence by age and sex

```{r fig1}
adults |>
  filter(!is.na(AgeDecade), !is.na(hypertension)) |>
  group_by(AgeDecade, Gender) |>
  summarise(prev = mean(hypertension), .groups = "drop") |>
  ggplot(aes(AgeDecade, prev, fill = Gender)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Age (years)", y = "Prevalence", fill = "Sex") +
  theme_minimal()
```

### Figure 2. Obesity prevalence by age and sex

```{r fig2}
adults |>
  filter(!is.na(AgeDecade), !is.na(obese)) |>
  group_by(AgeDecade, Gender) |>
  summarise(prev = mean(obese), .groups = "drop") |>
  ggplot(aes(AgeDecade, prev, fill = Gender)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Age (years)", y = "Prevalence", fill = "Sex") +
  theme_minimal()
```

### Table 2. Adjusted odds ratios for diabetes

```{r model}
m <- glm(factor(Diabetes) ~ BMI + PhysActive + smoker + Age + Gender,
         data = adults, family = binomial)

or_bmi <- tidy(m, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term == "BMI")

tbl_regression(m, exponentiate = TRUE) |> bold_p()
```

After adjustment for physical activity, smoking, age and sex, each one-unit
increase in BMI was associated with `r round(or_bmi$estimate, 2)` times the odds
of diabetes (95% CI `r round(or_bmi$conf.low, 2)` to `r round(or_bmi$conf.high, 2)`).
